---
title: "HW_434_14"
author: "Meghan Mokate"
date: "4/21/2021"
output:
  html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import libraries

```{r, echo = FALSE, include=FALSE}
library(tidyverse) # Our usual package
library("kableExtra") # Making cool tables in R
library("stevedata") # Some data from Steve...
```

## Load data
```{r}
data <- gss_wages
```

## Select only columns and rows that we need for our analysis

For our analysis, we only need the columns: year, realrinc, childs, and gender.

I also filter so that I only look at years 1998 through 2018 (given in two year increments).
```{r}
data_1 <- data %>%
  select(year, realrinc, childs, gender) %>%
  filter(year %in% (1998:2018)) 
```

## Select only columns and rows that we need for our analysis

Create columns for the following:

1. Average income for females
2. Average income for males
3. Number of children for females
4. Number of children for males
5. Difference in income
6. Difference in number of children

I also order these rows to group male, female, and difference
```{r, warning=FALSE}
table <- data_1 %>%
  group_by(year, gender) %>%
  summarize(avg_inc = mean(realrinc, na.rm = T),
         avg_child = mean (childs, na.rm = T)) %>%
  pivot_wider(names_from = gender,
              values_from = c(avg_inc, avg_child)) %>%
  mutate(inc_diff = avg_inc_Male - avg_inc_Female,
         child_diff = avg_child_Male - avg_child_Female) %>%
  select(year, 
         avg_inc_Male, avg_child_Male, 
         avg_inc_Female, avg_child_Female,
         inc_diff, child_diff)
table
```

## Format Table

I make the following changes to the table:

1. Change year to character
2. Create column for income change (to work with later)
3. Change the column names, with Year including a note (referenced later)
4. Right align all columns
5. Give the table a title
6. Specify digits for each column value
7. Add commas to numbers
8. Change font size and style
9. Make the column names have a light gray background
10. Give the syle a "hover" type
11. Add a header above to group columns for male, female, and the difference between the two for each group's income and number of children
12. Create a before and after crisis divider for the rows, with after crisis including a footnote
13. Color grouped columns to demonstrate they are separate
14. Italicize 2008 data since it stands out with big numbers and big differences
15. Change biggest difference value colors to red and smallest to green
16. Bold year values column
17. Add in footnotes
18. Add in pictures for increase / decrease

```{r, warning=FALSE}
t_format <- table %>%
  mutate(year = as.character(year)) %>%
  mutate("Income Change" = " ") %>%
  kbl(col.names = c("Year[note]", "Income", "Children",
                      "Income", "Children",
                      "Income", "Children", "Income Change"),
    align = "rrrrrrr",
    caption = "Average Income and # Children by by Gender 1998-2018",
    digits = c(0, 2, 2, 2, 2, 2, 2),
    format.args = list(big.mark = ",")) %>%
    kable_styling(font_size = 20,
                  html_font = "Arial",
                  full_width = F) %>%
  row_spec(0, bold = T, background = "#778899") %>%
  kable_material(c("hover")) %>%
  add_header_above(c(" " = 1, "Male" = 2,
                     "Female" = 2, "Difference" = 2, " " = 1)) %>% 
  group_rows("Before crisis", 1, 6, label_row_css = "background-color: #C0C0C0	") %>% 
  column_spec(2, background = c(rep("black", 0), "aliceblue", rep("black", 0)), rep(F, 0)) %>%
  column_spec(3, background = c(rep("black", 0), "aliceblue", rep("black", 0)), rep(F, 0)) %>%
  column_spec(4, background = c(rep("black", 0), "mistyrose", rep("black", 0)), rep(F, 0)) %>%
  column_spec(5, background = c(rep("black", 0), "mistyrose", rep("black", 0)), rep(F, 0)) %>%
  column_spec(6, background = c(rep("black", 0), "papayawhip", rep("black", 0)), rep(F, 0)) %>%
  column_spec(7, background = c(rep("black", 0), "papayawhip", rep("black", 0)), rep(F, 0)) %>%
  row_spec(6, italic = T) %>%
  column_spec(6, color = c(rep("black", 5), "red", "green", rep("black", 6))) %>%
  column_spec(7, color = c(rep("black", 6), "green", "red", rep("black", 8))) %>%
  column_spec(1, bold = T) %>%
  group_rows("After crisis[note]", 7, 11, label_row_css = "background-color: #C0C0C0") %>%
  add_footnote(c("Only includes 1998 through 2018", "We consider the end of crisis after 2009")) %>%
  column_spec(8, image = c('/Users/meghanmokate/Desktop/Spring%2021/both_dirs.png', 
                           '/Users/meghanmokate/Desktop/Spring%2021/up_arrow.png',
                           '/Users/meghanmokate/Desktop/Spring%2021/up_arrow.png',
                           "/Users/meghanmokate/Desktop/Spring%2021/down_arrow.png",
                           "/Users/meghanmokate/Desktop/Spring%2021/down_arrow.png",
                           "/Users/meghanmokate/Desktop/Spring%2021/up_arrow.png",
                           "/Users/meghanmokate/Desktop/Spring%2021/down_arrow.png",
                           "/Users/meghanmokate/Desktop/Spring%2021/up_arrow.png",
                           "/Users/meghanmokate/Desktop/Spring%2021/down_arrow.png",
                           "/Users/meghanmokate/Desktop/Spring%2021/down_arrow.png",
                           '/Users/meghanmokate/Desktop/Spring%2021/up_arrow.png'))
t_format
```

