---
title: "434_HW_12"
author: "Meghan Mokate"
date: "4/9/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load necessary packages

```{r, include=FALSE}
library(tidyverse)
```

## Load data and view details

```{r}
class(state.name)
class(sentences)

length(state.name)
length(sentences)
```

# Part 1

## 1. Extract the following states: Arizona, California, Illinois, Oregon

```{r}
str_subset(state.name, "Arizona|California|Illinois|Oregon") 
# use the | as an 'or' reference for each of the four states that I want to subset
```

## 2. Extract all states that begin with the letter a. Use both str_subset() and str_extract() to see how the syntax differs. In both cases, extract the full name of all the states.

```{r}
str_subset(state.name, "^A") 
# will return all values that begin with the letter A

str_extract(state.name, "^A.*")
# without the '.*' included, will only return the 
#beginning A and not the rest of the string value. Adding these in tells str to 
#return the entire string
```

## 3. Extract all states whose names finish with a OR e.

```{r}
str_subset(state.name, "[ae]$") 
# search for a or e then use $ to identify that these need 
#to be the last values of the string
```

## 4. Count how many state names are composed of two words (e.g. North Dakota)

```{r}
str_subset(state.name, "\\s") 
# shows the state names that are two words by identifying spaces
sum(str_count(state.name, "\\s")) 
# 10 states
# can use this since states with two words will have one space and with none will 
# have 0 spaces, so a sum of the count of spaces is effective
```

##5. Extract all states that contain at least one of these letters: n, t, w, c at any position within the name

```{r}
str_subset(state.name, "[ntwcNTWC]") 
# check for each of these letters
# include capital letters to ensure states that have the letters capitalized are included
```

## 6. Count how many states do not contain any one of these letters at any position: w, z, y

```{r}
sum(str_detect(state.name, "[WZYwzy]", negate = TRUE)) 
# check for each of these letters
# include capital letters to ensure states that don't have the letters capitalized are included
# negate specifies that I want the ones who do not contain

# 35 states
```

## 7. Extract all states that contain at least one of these letters: c or i. Exclude the initial letters from your count (e.g., do not include Colorado).

```{r}
str_subset(state.name, "[ci]") 
# check for each of these letters, excluding capitals
```

## 8. Identify which states are exactly 6-letter long (DO NOT use str_length here!!) AND states that are at least 6-letter long

```{r}
str_subset(state.name, "^.{6}$") 
# identify beginning of string, then a wildcard for 6 values long, then then end of the string

str_subset(state.name, ".{6,}") 
# look for states that are at least 6 letters long using {}
```

## 9. Create a regular expression that will match any string in state.name vector

```{r}
str_subset(state.name, "[A-Z][a-z]")
# all strings begin with a capital letter then have a lower case letter
```

## 10. Find all states that start with two consonants.

```{r}
str_subset(state.name, "^[BCDFGHJKLMNPQRSTVWXZ][bcdfghjklmnpqrstvwxz]") 
# start with the ^ to indicate the start, then reference capital consonants as
# the first letter, then lower case consonants as the second letter
```

## 11. Find all states that have two or more vowels in a row.

```{r}
str_subset(state.name, "[AEIOUaeiou][aeiou]") 
# for the first group, allow for capital or lower case vowels to be identified 
# in case a value had two vowels with one being the first letter
# then for the second group, we only have to worry about lower case
```

## 12. List the states that start with a vowel and end with a consonant

```{r}
str_subset(state.name, "^[AEIOU].*[bcdfghjklmnpqrstvwxz]$") 
# all states start with capital letters, so reference capital vowels using ^ to 
# indicate the beginning
# use the .* to separate beginning and end, allowing for any values between
# all states end with lowercase letters, so reference lowercase consonants 
#using $ to indicate the end
```

# Part 2: Create a few new columns to store the following information. Use dplyr functions, not the dollar sign framework.

## Load necessary data

```{r}
class(sentences)
length(sentences)

sentences_tbl <- as_tibble(sentences)
sentences_tbl
```

## 1. The number of letters in each sentence. What is the average number of letters across sentences?

```{r}
sentences_tbl_1 <- sentences_tbl %>%
  mutate(letter_count = str_count(value, "[A-Za-z]"))
# specify that the str_count should only count for letters

sentences_tbl_1

sentences_tbl_1 %>%
  summarise(avg_count = mean(letter_count)) 
# create a new column to find the average number of letters across sentences 
# (not added to the dataset)
# 31.32083 letters

```

## 2. A dummy variable equal to 1 if the sentence contains a word ending with “ing”, and 0 otherwise. Note that words not sentences should end in ing! How many words with ing in the dataset?

```{r}
sentences_tbl_2 <- sentences_tbl_1 %>%
  mutate(end_ing = ifelse(str_detect(value, "ing\\b"), 1, 0))
# looks for words that end in ing by including \\b after the word

sentences_tbl_2

sentences_tbl_2 %>%
  mutate(count_ing_end = str_count(value, "ing\\b")) %>%
  summarise(count = sum(count_ing_end)) 
# create a column to count the words ending in ing, and then count these 
# (not added to dataset)

# There are 53 words with that end with ing in the dataset

sentences_tbl_2 %>%
  mutate(count_ing = str_count(value, "ing")) %>%
  summarise(count = sum(count_ing))
# create a column to count the words that contain ing, and then count these 
# (not added to dataset)

# There are 64 words with that contain ing in the dataset
```

## 3. A dummy variable equal to 1 if the sentence contain a color such as red, orange, yellow, green, blue, purple, pink and 0 otherwise. Take advantage of the paste0 function to answer this question (see slides). Pay attention not to have a too permissive pattern! How many sentences contain colors?

```{r}
color <- c(" red ", " red.", "Red ", " orange ", " orange.", "Orange ", " yellow ", " yellow.", "Yellow ",
          " green ", " green.", "Green ", " blue ", " blue.", "Blue ", " purple ", " purple.", "Purple ",
          " pink ", " pink.", "Pink ", " gold ", " gold.", "Gold ", " white ", " white.", "White ",
          " black ", " black.", "Black")

# identify every possible combination of colors that may appear
# I specify that the colors can either have a space after them / be capitalized 
# with a space after them for the middle or end of a sentence. 
# Or, they can have a space before and a period for the end of the sentence
#This prevents words containing the color words (ex: ordered contains 'red') from being included

search <- paste0(color, collapse = "|")
# use a paste0 with collapse as | as to indicate 'or' for all colors 
 
sentences_tbl_3 <- sentences_tbl_2 %>%
  mutate(color = ifelse(str_detect(value, search), 1, 0))
# use ifelse with a str_detect to search for the color words in the value column
# return a 1 if they are included and a 0 if nor in a new column, color

sentences_tbl_3

sentences_tbl_3 %>%
  summarise(count = sum(color))
# create a column to count the sentences that contain color words 
# (not added to dataset)

# 53 sentences
```

## 4. The number of words in each sentence. This requires a bit of creativity and there are likely multiple solutions! Give it a fair try but it won’t negatively affect your grade if you cannot figure it out :)

```{r}
sentences_tbl_4 <- sentences_tbl_3 %>%
  mutate(word_count = sapply(strsplit(value, " "), length))

# use strsplit on the value column with a split of " " and a function of length 
# within a sapply to count the amount of words in each sentences 
# add it to a new column called word_count

sentences_tbl_4
```


