---
title: "434 Assignment 9"
author: "Meghan Mokate"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document: 
    toc: true
    toc_depth: 2
    toc_float: yes
    theme: simplex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set up the workplace
### Upload needed packages
```{r}
library(tidyverse)
library(ggalt)
library(forcats)
```

### Import Data 
```{r}
crimes_2010 <- read_csv("Crimes_2010.csv")
crimes_2019 <- read_csv("Crimes_2019.csv")
```

# Clean and Organize Data
### Remove Categories we don't need for both datasets and view them
```{r}
narcotics_2010_r <-
  crimes_2010 %>%
  filter(`Primary Type` == "NARCOTICS") %>%
  select(c("Date",
           "Description",
           "Ward"))

narcotics_2010_r

narcotics_2019_r <-
  crimes_2019 %>%
  filter(`Primary Type` == "NARCOTICS") %>%
  select(c("Date",
           "Description",
           "Ward"))

narcotics_2019_r
```

### Combine datasets
```{r}
narcotics_2010_2019 <-
  full_join(narcotics_2010_r, narcotics_2019_r)

narcotics_2010_2019
```

### Edit the class of the date column to be the year
```{r}
narcotics_2010_2019 <-
  narcotics_2010_2019 %>%
  separate(Date,
           into = c("Month", "Day", "Year")) %>%
  select(Year, Description, Ward)

narcotics_2010_2019
```

### Since some of these dates are in 2020, filter them out.
```{r}
narcotics_2010_2019$Year <- as.numeric(narcotics_2010_2019$Year)
class(narcotics_2010_2019$Year)

narcotics_2010_2019 <-
  narcotics_2010_2019 %>%
  filter(Year != 2020)

narcotics_2010_2019
```

## Code Type Of Narcotics
### See what drug types the descriptions can be grouped into
```{r}
unique(narcotics_2010_2019$Description)
```

Create new dataset with a new column that labels drug types more broadly for analyzing at the group level and use str_detect to search for certain words that identify the drug
```{r}
narcotics_2010_2019_1 <- narcotics_2010_2019 %>%
  mutate(drug_type = ifelse(str_detect(`Description`, 'CANNABIS'), "CANNABIS",`Description`)) %>%
  mutate(drug_type = ifelse(str_detect(`Description`, 'HEROIN'), "HEROIN", drug_type)) %>%
  mutate(drug_type = ifelse(str_detect(`Description`, 'CRACK'), "CRACK", drug_type)) %>%
  mutate(drug_type = ifelse(str_detect(`Description`, 'PCP'), "PCP", drug_type)) %>%
  mutate(drug_type = ifelse(str_detect(`Description`, 'PARAPHERNALIA'), "PARAPHERNALIA", drug_type)) %>%
  mutate(drug_type = ifelse(str_detect(`Description`, 'BARBITURATE'), "BARBITURATES", drug_type)) %>% 
  mutate(drug_type = ifelse(str_detect(`Description`, 'NARCOTICS'), "OTHER NARCOTICS", drug_type)) %>%
  mutate(drug_type = ifelse(str_detect(`Description`, 'HALLUCINOGEN'), "HALLUCINOGENS", drug_type)) %>%
  mutate(drug_type = ifelse(str_detect(`Description`, 'COCAINE'), "COCAINE", drug_type)) %>%
  mutate(drug_type = ifelse(str_detect(`Description`, 'AMPHETAMINES'), "AMPHETAMINES", drug_type)) %>%
  mutate(drug_type = ifelse(str_detect(`Description`, 'METHAMPHETAMINE'), "METHAMPHETAMINE", drug_type)) %>%
  mutate(drug_type = ifelse(str_detect(`Description`, 'SYNTHETIC DRUGS'), 'SYNTHETIC DRUGS', drug_type)) %>%
  mutate(drug_type = ifelse(str_detect(`Description`, 'DRUG EQUIPMENT'), 'DRUG EQUIPMENT', drug_type)) %>%
  mutate(drug_type = ifelse(str_detect(`Description`, 'FORGE PRESCRIPTION'), 'ALTER/FORGE PRESCRIPTION', drug_type)) %>%
  mutate(drug_type = ifelse(str_detect(`Description`, 'FORFEIT PROPERTY'), 'PROPERTY FORFEITURE', drug_type)) %>%
  mutate(drug_type = ifelse(str_detect(`Description`, 'SYNTHETIC MARIJUANA'), 'SYNTHETIC MARIJUANA', drug_type)) %>%
  mutate(drug_type = ifelse(str_detect(`Description`, 'CRIMINAL DRUG CONSPIRACY'), "OTHER DRUG CRIMES", drug_type)) %>%
  mutate(drug_type = ifelse(str_detect(`Description`, 'CONT SUBS:FAIL TO MAINT RECORD'), "OTHER DRUG CRIMES", drug_type)) %>%
  mutate(drug_type = ifelse(str_detect(`Description`, 'HYPODERMIC NEEDLE'), 'HYPODERMIC NEEDLE', drug_type)) %>%
  mutate(drug_type = ifelse(str_detect(`Description`, 'BARBITUATES'), "BARBITUATES", drug_type)) %>%
  mutate(drug_type = ifelse(str_detect(`Description`, 'LOOK-ALIKE'), "OTHER NARCOTICS", drug_type)) %>%
  mutate(drug_type = ifelse(str_detect(`Description`, '<18'), "OTHER DRUG CRIMES", drug_type))
```

### Check on the new column groupings to make sure everything is grouped and makes sense
```{r}
unique(narcotics_2010_2019_1$drug_type)
```

# Analyze Data
### Take a quick look at frequency of reported crimes for each drug type
```{r}
ggplot(data = narcotics_2010_2019_1) +
  geom_bar(mapping = aes(y = drug_type))
```

### I want a point for 2010 and for 2019. I need the data to have a column for counts from each year.
```{r}
narcotics_a <-
  narcotics_2010_2019_1 %>% 
  filter(drug_type != "SYNTHETIC MARIJUANA") %>%
  group_by(drug_type, Year) %>%
  count(drug_type, name = "Reports")

narcotics_a

narcotics <- narcotics_a %>%
  pivot_wider(names_from = "Year",
              values_from = "Reports") %>%
  mutate(per_change = round(((`2019` - `2010`)/`2010`) * 100, digits = 2)) %>%
  mutate(per_change_ordering = abs(round(((`2019` - `2010`)/`2010`) * 100, digits = 2))) %>%
  arrange(desc(per_change_ordering)) %>%
  mutate(drug_type = as.factor(drug_type))
  
  
narcotics$drug_type <-  fct_reorder(.f = narcotics$drug_type, .x = narcotics$per_change_ordering, .desc = F)
```

### Take a look at the dataset
```{r}
narcotics
```

# Plot Graph #1
### Plot the data with the two points
```{r}
ggplot() +
  geom_dumbbell(data = narcotics, aes(y = drug_type, x = `2010`, xend = `2019`),
                size = 1.0, color = "black", size_x = 3, size_xend = 3,
                colour_x = "#27A192", colour_xend = "#A12736",
                alpha = 0.5) + #Use this to change the transparency of the dots
  labs(title = "Change in Narcotic Crime Reports from 2010 to 2019",
       x = "Number of Reports",
       y = "Narcotic Crime Reported") 
```

## Enhancing the plot 
From the base plot, we need to add a number of elements for clarity and for practical purposes. First, we need to duplicate the plots of the dumbbell so we have a mapping that we can create a legend for. This way we can have a legend for the dumbbells and make it clear which color corresponds to what year.

A second element that should be added is something to convey the *scale of the change*, 
whether in absolute terms or % change. This is especially important because the initial plot indicates we might want to transform the scale of the x-axis, as the large change in cannabis 
numbers is making it hard to view the changes in the other drug categories that have lower counts to begin with. One possible solution may be to use a log transformation, you just need to be sure this is clearly conveyed since log scales are not commonly used. 
 
Since I plan on including the % difference as a geom_text, I also plan on using the 
reordered factor so I can display the drug_types in descending order of the absolute value
of the % change so the most extreme changes (both positive and negative) display first. 

For the most part, this plot reads like most other dumbbell plots, it shows the 
change in number of reports in 2010 (red) and in 2019 (blue) with the crime types
listed in descending order of the abs() of the % change, so the crime types are
listed in descending order of the total % change with the greatest changes (in either direction) 
listed first. Special attention should be given to the fact that **the x-axis of this plot uses a logged scale**, meaning that visually equal areas do not represent equal
actual values. This means that lines towards the left of the plot will appear longer
even if the absolute change in reported crimes is fairly small. For example, the
line for cannabis is shorter than the line for "Other Drug Crimes" but represents a
greater absolute change of nearly 20,000, despite the change in other crimes being 
a difference of only 50.  
```{r, fig.width=12,fig.height=11}
citywide_plot <- ggplot() +
  geom_dumbbell(data = narcotics, aes(y = factor(drug_type), x = `2010`, xend = `2019`),
                size = 1.0, color = "black", size_x = 3, size_xend = 3,
                colour_x = "#FF0000", colour_xend = "41b7e6",
                alpha = 0.5) +
  geom_point(data = narcotics_a, aes(x = Reports, y = drug_type, color = as.factor(Year)), size = 3, alpha = 0.5) +
  geom_rect(data = narcotics, aes(xmin = 50000, xmax = Inf, ymin = -Inf, ymax = Inf), fill = "#FFFFFF") +
  geom_text(data = narcotics, aes(label = paste0(per_change, "%"), y = drug_type, x = 50000), color = "#41b7e6", fontface = "bold", size = 4) +
  geom_text(data = filter(narcotics, drug_type == "METHAMPHETAMINE"), aes(x = 50000, y = drug_type, label = "% Difference"), color = "#41b7e6", size = 4.1, hjust = .6, vjust = -1.2, fontface = "bold") +
 labs(title = "Change in Narcotic Crime Reports in Chicago, IL from 2010 to 2019",
      subtitle = "with percent change from 2010 to 2019, City-Wide",
      caption = "X-Axis has been transformed by log10 for illustative purposes. Items on the Y-Axis are arranged in descending order of the absolute value of the percent change from 2010 to 2019.", 
      x = "Number of Reports",
      y = "Narcotic Crime Reported") +
  scale_x_continuous(trans = 'log10') + 
  scale_color_manual(name = "", values = c("#FF0000", "#41b7e6")) +
  theme_minimal() +
  theme(plot.title = element_text(color = "#FF0000", size = 20, face = "bold"),
        plot.subtitle = element_text(face = "bold.italic"),
        plot.caption = element_text(face = "italic", hjust = 0),
        legend.position = "bottom") 

citywide_plot

ggsave(filename = "Citywide Plot.png",
       plot = citywide_plot,
       width = 10,
       height = 10,
       units = "in")  
```

# Analysis of Changes by Ward
### See which 5 wards saw the smallest and largest percent changes in narcotics reports from 2010 to 2019

We can see that wards 24, 36, 27, 37 and 22 saw the *smallest* changes in narcotics crimes reported, though all were > 55%.

We can see that wards 39, 15, 38, 46, and 2 saw the *largest* changes in narcotics crimes reported. 
```{r}
narcotics_b <-
  narcotics_2010_2019_1 %>%
  group_by(Ward, drug_type, Year) %>%
  count(drug_type, name = "Reports")

narcotics_b

ward_narcotics <-
  narcotics_b %>%
  group_by(Ward) %>%
  summarise(report_2010 = sum(Reports[Year == 2010], na.rm = T),
            report_2019 = sum(Reports[Year == 2019], na.rm = T),
            diff = report_2019 - report_2010,
            perc_change = (diff / report_2010) * 100) %>%
  arrange(desc(perc_change))
ward_narcotics

ward_narcotics %>%
  top_n(5, wt = perc_change)

ward_narcotics %>%
  top_n(5, wt = perc_change)

ward_narcotics %>%
  top_n(-5, wt = perc_change)
```

### See which 5 wards had the highest numbers of narcotic crime reports

Wards 28, 24, 27, 37, and 16 had the *highest* number of reports.
We can see that 3 (24, 27, 37) of the 5 wards with the highest number of narcotic crime reports in 2010 saw the lowest percent changes in narcotics crimes in 2019.

24 = North Lawndale, some West Garfield Park
27 = West Loop, parts of East Garfield and Humboldt Parks
37 = Parts of Austin and Humboldt Park
```{r}
ward_narcotics %>%
  arrange(desc(perc_change))
```

# Plot Graph #2
Knowing that wards 24, 36, 27, 37, and 22 saw the least overall change in drug crime and three of them were in the highest number of crime reports in 2010, let's look at the same plot we had before, but instead of city-wide crime data, let's only look at these specific wards to see if any city-wide trends change in scale or in direction.

From this we can see that most of the trends hold true, **synthetic drugs, barbituates, and meth** all 
increased at both levels, BUT synthetic drugs saw a MUCH greater increase in these
wards then city-wide (300% increase v. 91% increase). PCP crimes also saw a clear *increase* in these
wards unlike the generally unchanged city-wide rate.  

```{r, fig.width=12,fig.height=11}
narcotics_c <-  narcotics_2010_2019_1 %>%
  filter(drug_type != "SYNTHETIC MARIJUANA") %>%
  filter(Ward == 24 | Ward == 36 | Ward == 27 | Ward == 37| Ward == 22) %>%
  group_by(drug_type, Year) %>%
  count(drug_type, name = "Reports")

narcotics_wards <- narcotics_c %>%
  pivot_wider(names_from = "Year",
              values_from = "Reports") %>%
  mutate(per_change = round(((`2019` - `2010`)/`2010`) * 100, digits = 2)) %>%
  mutate(per_change_ordering = abs(round(((`2019` - `2010`)/`2010`) * 100, digits = 2))) %>%
  arrange(desc(per_change_ordering)) %>%
  mutate(drug_type = as.factor(drug_type))

narcotics_wards$drug_type <-  fct_reorder(.f = narcotics_wards$drug_type, .x = narcotics_wards$per_change_ordering, .desc = F)

wards_plot <- ggplot() +
  geom_dumbbell(data = narcotics_wards, aes(y = drug_type, x = `2010`, xend = `2019`),
                size = 1.0, color = "black", size_x = 3, size_xend = 3,
                colour_x = "#FF0000", colour_xend = "41b7e6",
                alpha = 0.5) + 
  geom_point(data = narcotics_c, aes(x = Reports, y = drug_type, color = as.factor(Year)), size = 3, alpha = 0.5) +
  geom_rect(data = narcotics_wards, aes(xmin = 50000, xmax = Inf, ymin = -Inf, ymax = Inf), fill = "#FFFFFF") +
  geom_text(data = narcotics_wards, aes(label = paste0(per_change, "%"), y = drug_type, x = 50000), color = "#FF0000", fontface = "bold", size = 4) +
  geom_text(data = filter(narcotics_wards, drug_type == 'SYNTHETIC DRUGS'), aes(x = 50000, y = drug_type, label = "% Difference"), color = "#FF0000", size = 4.1, hjust = .6, vjust = -1.2, fontface = "bold") +
  labs(title = "Change in Narcotic Crime Reports in Chicago, IL from 2010 to 2019",
       subtitle = "with percent change from 2010 to 2019, Top 5 Wards with lowest percent change (Wards 24, 36, 27, 37, 22)",
       caption = "X-Axis has been transformed by log10 for illustative purposes. Items on the Y-Axis are arranged in descending order of the absolute value of the percent change from 2010 to 2019.", 
       x = "Number of Reports",
       y = "Narcotic Crime Reported") +
  scale_x_continuous(trans = 'log10') + 
  scale_color_manual(name = "", values = c("#FF0000", "#41b7e6")) +
  theme_minimal() +
  theme(plot.title = element_text(color = "#41b7e6", size = 20, face = "bold"),
        plot.subtitle = element_text(face = "bold.italic"),
        plot.caption = element_text(face = "italic", hjust = 1.5),
        legend.position = "bottom")

wards_plot

ggsave(filename = "Wards Plot.png",
       plot = wards_plot,
       width = 10,
       height = 10,
       units = "in")    
```

